Peptide retention time prediction

### 2015-01-13 retention.tex
Edits

The amount of time that a peptide needs to travel through the column
is called retention time, shortly $t_R$. One often uses the term
retention time to refer to the time corresponding to the apex of each
peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). Typically,
the more hydrophobic the peptide, the longer the retention time.

@@

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
ANN & Link? & \cite{petritis2003,petritis2006improved} \\
RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\

@@

The idea of identifying proteins based on entire peptides, rather than
peptide fragments, is not new. In fact, this was the principle
underlying peptide mass fingerprinting, one of the first methods used
to identify proteins via mass spectrometry. However, when introduced
in the 90's, mass fingerprinting could not be applied to unequivocally
identify proteins in complex biological samples. Consequently, shotgun
proteomics has gained popularity instead. Nevertheless, with great
developments in instrumentation leading to sub-ppm mass accuracies,
the possibility of using mass fingerprinting-like methods came back to
attention~\cite{iamt1}. Our study showed how improved mass accuracies
and retention time predictions enable the use of unfragmented ions
even in the context of complex peptide mixtures.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered bolzman machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that the performance of predictors is often enhanced if
one can structure the prediction into multiple layers. The drawback of
such models are that they require large training sets. Given the very
large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM), are dependent on the predictability of a
peptide's behavior. In such experiments the retention times and the
suitable transitions are needed as input to the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiments
would lead to a more direct experimental design and hopefully to more
comprehensie coverages. Thus {\em de novo} retention time predictions
have an important role to fill.

@@

%%  LocalWords:  hydrophobicity chromatography elute peptides physico
%%  LocalWords:  electrospray spectrometry analytical proteomics RPLC
%%  LocalWords:  posttranslational phosphorylation proteome analytes
%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls
%%  LocalWords:  workflows trypsin hydrophobicities chromatographic

>>>>

The amount of time that a peptide needs to travel through the column
is called retention time, shortly $t_R$. One often uses the term
retention time to refer to the time corresponding to the apex of each
peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). This time
point will be proportional to the peptide's ability to lower the
surface potential between the beads and the solvent, a property that
we normally refer to as the peptide's hydrophobicity.  Typically, the
more hydrophobic the peptide, the longer the retention time.

@@

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
% ANN & Link? & \cite{petritis2003,petritis2006improved} \\
RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\

@@

The idea of identifying proteins based on entire peptides, rather than
peptide fragments, is not new. In fact, this was the principle
underlying peptide mass fingerprinting, one of the first methods used
to identify proteins via mass spectrometry. However, when introduced
in the 90's, mass fingerprinting could not be applied to unequivocally
identify proteins in complex biological samples. Consequently, shotgun
proteomics has gained popularity instead. Nevertheless, with great
developments in instrumentation leading to sub-ppm mass accuracy,
the possibility of using mass fingerprinting-like methods came back to
attention~\cite{iamt1}. Our study showed how improved mass accuracy
and retention time predictions enable the use of unfragmented ions
even in the context of complex peptide mixtures.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as bigdata, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered Boltzmann machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that the performance of predictors is often enhanced if
one can structure the prediction into multiple layers. The drawback of
such models are that they require large training sets. Given the very
large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM), are dependent on the predictability of a
peptide's behavior. In such experiments the retention times and the
suitable transitions are needed as input to the analysis. Frequently,
such information is derived from previous shotgun analysis. As the
capacity of targeted proteomics experiments expands by the
introduction of the more comprehensive data independent acquisition
techniques \cite{Venable2004} (a.k.a. SWATH analysis), such
pre-identification of peptide transitions and retention times
threatens to become a limiting factor. Our current methods to identify
peptides from DIA are only as comprehensive as the previous shotgun
analysis would be. An attractive alternative is to predict retention
time and transition using {\em de novo} predictions from the peptides
amino acid sequences.  Such a system of identifications not dependent
on previous peptide identifications from shotgun proteomics experiments
would lead to a more direct experimental design and hopefully to more
comprehensive coverage. Thus {\em de novo} retention time predictions
have an important role to fill.

@@

%%  LocalWords:  hydrophobicity chromatography elute peptides physico
%%  LocalWords:  electrospray spectrometry analytical proteomics RPLC
%%  LocalWords:  posttranslational phosphorylation proteome analytes
%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls et
%%  LocalWords:  workflows trypsin hydrophobicities chromatographic
%%  LocalWords:  alkyl peptide's elution SSRCalc BioLCCC RTModel iRT
%%  LocalWords:  RTPredict Krokhin al QSRR Klammer SVMs SVM Moruz pre
%%  LocalWords:  dataset workflow outliers posttranslationally silico
%%  LocalWords:  phosphorylated artifactual phosphopeptides COmbined
%%  LocalWords:  FRActional DIagonal COFRADIC datasets Palmblad
%%  LocalWords:  Strittmatter unfragmented
